---
title: "Introduction to PeCorA"
author: "Maria Dermit"
date: "`r BiocStyle::doc_date()`"
package: "`r pkg_ver('PeCorA')`"
output: 
    BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{PeCorA Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
  %\VignetteEncoding{UTF-8}
bibliography: ["PeCorA.bib"]
---

\pagebreak

# Abstract

`PeCorA` (peptide correlation analysis) is a package for detection of quantitative disagreements between peptides mapped to the same protein. 
This package provides an integrated analysis workflow for LFQ data that requires tabular input (e.g. peptide.txt file) as generated by quantitative analysis software of raw mass spectrometry data, such as MaxQuant [@Cox2014].
Functions are provided for data preparation, scaling, linear modeling and statistical testing of deferentially discordant peptides.
It also includes tools for making a table of the peptides that disagree. 
Finally, visualization tools are provided to explore the results, including boxplot representations. 


# Installation and loading

Start R and install the PeCorA package:
```{r install, eval = FALSE}
if (!requireNamespace("BiocManager", quietly=TRUE))
    install.packages("BiocManager")
BiocManager::install("PeCorA")
library("PeCorA")
```

Once you have the package installed, load `PeCorA` into R.

```{r setup, message=FALSE}
library(PeCorA)
```

# Converting MaxQuant peptide output into PeCorA format

## Example dataset: COVID-19 plasma proteomics

We analyze a large-scale analysis of COVID19 severity. The data set was composed
of over 100 plasma samples from three groups: (1) COVID-19-driven acute
respiratory distress syndrome (ARDS) patients, (2) non-COVID-19-driven ARDS
patients, and (3) pooled plasma control sample extracted with each batch as 
quality control [@Overmyer2020].
The raw mass spectrometry data were first analyzed using MaxQuant
(version 1.6.10.43)[@Cox2014] and the resulting "peptides.txt" file is used as input for the downstream analysis.
We filtered non-relevant information (e.g. amino acid count) and provide this dataset
with PeCorA package 

## Loading of the data

```{r load-data}
data(peptides_data_filtered)
```

This dataset has the following dimensions:
```{r dim-data}
dim(peptides_data_filtered)
```

The "Leading.razor.protein", "Sequence" and "LFQ.intensity" columns will be used 
for subsequent analysis. There are three biological conditions in this 
experiment,and the names captured in the LFQ.intensity variables will be used
for comparisons

## Data preparation

You can prepare MaxQuant peptide output into PeCorA-ready format using the 
function `import_LFQ_PeCorA`.

```{r import-LFQ, message=FALSE}
pecora_format <- import_LFQ_PeCorA(peptides_data_filtered,
                                   protein = 'Leading.razor.protein',
                                   sequence='Sequence',
                                   condition1='control',
                                   condition2='_COVID',
                                   condition3='NON.COVID')
                                    
```

# Scaling and centering peptides

`PeCorA_preprocessing` initially filters the values to include only precursors
with measured MS1 areas in all samples. Next, the peak areas are log2 
transformed, and the global distribution of all peak areas was scaled to have the
same center. Finally, each peptide is center relative to the mean of the control
group’s peak area. 

```{r preprocessing, message=FALSE}
scaled_peptides <- PeCorA_preprocessing(pecora_format,
                                        area_column_name=4,
                                         threshold_to_filter=100,
                                         control_name="control")
```


# Running PeCorA analysis

`PeCorA` loops through proteins with >2 peptides, and records a linear model on 
the peptide precursors for each of those protein recording a adjust p-value 
within each protein. It makes a dataframe with the peptides that disagree, 
sorting smaller adj_pval values at the top of table.

```{r PeCorA-core, message=FALSE}
disagree_peptides <-  PeCorA(scaled_peptides)
```

# Plotting PeCorA results

Example boxplot of a significant peptide detected in pro-thrombine in the
COVID-19 plasma proteomics dataset as in [@Dermit2020].
```{r plotting}
PeCorA_plotting_plot<-PeCorA_plotting(disagree_peptides,
                                      disagree_peptides[12,],
                                      scaled_peptides)
PeCorA_plotting_plot
```


# Session information

``` {r session_info, echo = FALSE}
sessionInfo()
```

# References {-}


